### 2019-11-07 Paper_shortened_pdf.Rmd
Sent to Renske

abstract: |
  The question of whether fathers and mothers play similar roles in children’s
  emotional adjustment has been approached differently in different research
  traditions: Developmental psychologists have focused on qualitative parenting,
  and sociologists have focused on quantitative child care involvement. This
  longitudinal study (N = 1884, age 6-12) bridges these approaches, 
  investigating qualitative and quantitative involvement, and their interaction,
  in relation to emotional adjustment. On average, evidence for mothers'
  and fathers' unique roles was limited.
  When both parents were more involved, both engaged
  more in desirable qualitative parenting. When mothers were disproportionately
  involved, however, parenting roles were more differentiated:
  Fathers displayed less
  desirable qualitative parenting, and children were less well-adjusted.
  This suggests that all family members benefit from a more equal
  childcare distribution.

@@

\vspace{\baselineskip}
Emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of the quantitative amount of care parents provide (who does what), and in terms of qualitative parenting behaviors (how do parents behave towards children). Nevertheless, the existing literature has two major shortcomings. Firstly, quantitative involvement and qualitative parenting behaviors are rarely jointly investigated. Therefore, we currently don't know how the interplay between quantitative and qualitative involvement relates to emotional development. Secondly, although fathers have been theorized to play a unique role in emotional adjustment development, empirical work has rarely addressed the joint and relative contributions of fathers' and mothers' involvement. The present study addresses overcomes these limitations by incorporating both quantitative and quantitative measures of mothers' and fathers' involvement. We use sophisticated methods to examine how joint and relative quantitative involvement moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment.

@@

Attachment relationships are grounded in a balance between security and exploration [@grossmannDeveloppementAttachementAdaptation1998]. Fatherhood researchers have posited that warmth and security are more central to mother-child attachment relationships, whereas father-child relationships have been construed as "activation relationships", with greater emphasis on exploration [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers stimulate a broad spectrum of emotions, thereby expanding children's emotional repertoire, and engaging their developing regulatory abilities  [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000]. Consequently, fathers and mothers are theorized influence emotional adjustment development in somewhat different ways.

@@

It is now widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004]. Nevertheless, these dimensions are rarely investigated together [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

@@

### Parenting effects

Early developmental research focused on parental socialization; the notion that parents shape child development through proximal mechanisms, such as qualitative parenting behaviors [@maccobySocializationContextFamily1983; see also @vanlissaRoleFathersMothers2018a]. Qualitative parenting behaviors have been operationalized in terms of the dimensions underlying parenting styles: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Substantial evidence indicates that responsiveness, which involves warm and supportive parenting, is positively associated with children's socio-emotional adjustment [@khalequePerceivedParentalAcceptanceRejection2002; @morrisRoleFamilyContext2007].

The literature linking demandingness and emotional adjustment is less conclusive [@morrisRoleFamilyContext2007]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Some of these might help children meet societal expectations, whereas others are overly restrictive, and interfere with children's autonomy needs [@barberCentralityControlParenting2013]. We examined two aspects of demandingness: Consistency; the extent to which consequences for transgressions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. Consistency has been found to correlate strongly with emotional adjustment [@brandPerceivedParentalInconsistency1990]. Inductive reasoning is thought to promote the internalization of parental expectations, including display rules for emotions, and might thus play a role in emotional adjustment development as well [@grusecImpactParentalDiscipline1994].

Negative parenting is an additional qualitative parenting dimension that appears to be negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. We operationalized negative parenting as parental anger. Parental anger is thought to model anger in children, and undermine their emotion regulation [@scaramellaParentingChildrenDistress2008].

Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' influence [@flouriFathersBehaviorsChildren2010]. Although some research suggests that both parents influence children similarly [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique contributions of mothers and fathers. For various reasons, mothers and fathers sometimes appear to fulfill somewhat specialized roles [@bogelsDoesFatherKnow2011]. In line with the theoretical notions of "attachment" and "activation" relationships, mothers tend to encourage, and respond more supportively to, emotional expression, whereas fathers structure and sanction emotion expressions to a greater extent [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, mothers' supportive behavior was found to be a more salient socialization mechanism than control for adolescents' emotional adjustment, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. Similar differentiated associations were also found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is some evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is incongruous with the notion of the maternal "attachment" relationship. We might therefore expect emotional adjustment to be more strongly associated with maternal warmth, and with paternal demanding behaviors, such as consistency and inductive reasoning.

### Child effects

A view of the family as a dynamic system, whose members reciprocally influence one another, has become commonplace [@kerrPerceivedParentingStyle2012]. Nevertheless, bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. The available evidence suggests that maladjustment elicits less parental warmth, more demandingness, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004; @nepplIntergenerationalContinuityParenting2009; @lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional support that emotional maladjustment predicts decreased parental support, and greater control [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014; @wijsbroekDirectionEffectsPerceived2011]. Thus, parents' display punitive responses to maladjustment, including decreased support and increased negative parenting, and corrective responses, such as increased demandingness. We might therefore expect emotional adjustment to positively predict parental warmth, and negatively predict consistency, inductive reasoning, and anger.

### Direction of effects

@@

@@

Quantitative involvement has been researched extensively in terms of the absolute amount of childcare provided [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the **division** of childcare tasks is more relevant for children's adjustment [@krollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004; @craigHowMothersFathers2011; @feinbergInternalStructureEcological2003]. Indeed, the equity of the childcare distribution is associated with socio-emotional development [@deutschPaternalParticipationChild2001]. Moreover, an unequal distribution often implies that women perform a disproportionate share of the work [@craigDoesFatherCare2006]. This strains the marital relationship [@shockleyItNotWhat2018], and erodes parenting quality.

A potential confound might be introduced when operationalizing quantitative involvement in terms of both parents' absolute involvement: Effects that appears to be due to fathers' greater involvement might be due to children receiving **more** total care [see @deutschPaternalParticipationChild2001]. This confound can be mitigated by examining the effect of the distribution of childcare, while controlling for total care. Some prior research has attempted to achieve this by assessing the childcare distribution directly, by asking which parent is most responsible for certain tasks [@krollEarlyFatherInvolvement2016; @keizerInfluenceFathersMothers2019]. Alternatively, when quantitative involvement has been measured in terms of absolute involvement, one can estimate a latent mean and difference to statistically disentangle joint involvement from relative involvement [@mcardleLatentVariableModeling2009]. Joint involvement reflects the total amount of care provided to children, and relative involvement reflects the childcare distribution.

@@

The amount of time parents spend with children is associated with the quality of parenting behaviors  [@sabattiniRelationMothersFathers2004]. Indeed, more involved parents tend to experience greater parenting self-efficacy, possibly reflecting the effects of experience [@gialloMothersFathersInvolvement2013a]. Furthermore, when both parents are more involved, they monitor each other's parenting behaviors to ensure that children are raised well [@mclanahanConsequencesFatherAbsence2014]. Consequently, we might expect greater joint involvement to positively predict both parents' parenting quality.

With regard to differential involvement, we might also expect the more involved parent to display more desirable qualitative parenting behaviors, due to the effects of experience [@sabattiniRelationMothersFathers2004]. However, the benefit of parents monitoring each other's child-rearing should be diminished when parents are not equally involved [@mclanahanConsequencesFatherAbsence2014]. Indeed, retrospective reports indicate that an unequal childcare distribution is associated with suboptimal parenting behaviors: Increased authoritarian parenting in mothers, and increased disengaged parenting in fathers [@sabattiniRelationMothersFathers2004]. Additionally, greater differential involvement might undermine both parents' involvement quality, through its effect on the marital relationship. Childcare responsibilities are a heavy burden [@deutschPaternalParticipationChild2001]. An unfair distribution strains the interparental relationship, interferes with sensitive parenting, and undermines child well-being [@davies2006interparental]. When care is distributed relatively equally, mothers tend to be more satisfied [@pharesFathersMothersInvolvement2009], and the family system more harmonious. Consequently, balanced co-parenting is associated with greater parental warmth [@barnettInterdependenceParentingMothers2008]. We might thus expect greater differential involvement to be negatively associated with both parents' parenting quality, and with children's emotional adjustment.

@@

Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. This suggests that the total amount of care provided should be positively associated with children's emotional adjustment. Differential involvement, on the other hand, might be negatively linked to emotional adjustment: When children perceive that one of their parents is relatively uninvolved, they likely feel less supported than when both parents are equally invested in their well-being [@deutschPaternalParticipationChild2001].

@@

Interactions between involvement quality and quantity have largely been overlooked in theoretical models of father involvement; perhaps in part because these two dimensions have rarely been jointly investigated [@jaffeeLifeFatherBenefits2003]. In this respect, theories of father involvement can be complemented by Feinberg's coparenting model [@feinbergInternalStructureEcological2003]. Coparenting refers to the ways parents relate to each other in the role of parent. Although the coparenting literature is typically invoked to describe child-rearing in divorced families, it also holds insights for intact families. According to Feinberg, the "division of labor" in childcare, analogous to our operationalization of differential involvement, affects qualitative child-rearing behaviors and children's adjustment directly, and moderates the link between these two constructs.

@@

Sex differences are a relevant factor in the interplay between parenting and emotional adjustment. Parents tend to raise boys and girls differently; expressing more control with girls [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2018a], and more negative parenting with boys [@strausCorporalPunishmentAmerican1999]. Furthermore, social learning theory posits that the same-sex parent plays a stronger role in socialization; particularly for sex-typed behavior, such as emotion display rules [@busseySocialCognitiveTheory1999]. Finally, mounting evidence suggests that parental sensitivity is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2018a]. We therefore hypothesize that parents will display more responsiveness, structure, and inductive reasoning towards girls, and more negative parenting towards boys. Furthermore, we predict stronger associations between qualitative parenting and emotional adjustment between children and the same-sex parent.

Although sex differences in emotional adjustment are sometimes found in older children [@silkAdolescentsEmotionRegulation2003; @vanlissaRoleFathersMothers2018a], a meta-analysis found little evidence for such sex differences in young children [@else-questGenderDifferencesTemperament2006a]. Consequently, we have no hypothesis regarding sex differences in children's average levels of emotional adjustment.

@@

The present study aims to make several contributions. First, we include both mothers and fathers, in order to tease apart their common and unique roles in children's emotional adjustment development. Second, we incorporate both qualitative and quantitative measures of involvement. Third, we disentangle stable between-family differences from dynamic within-family predictive processes, in order to more accurately assess the direction of effects between qualitative involvement and emotional adjustment. Fourth, we examined interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work. Finally, we examined sex as a moderator of between-family differences in all study variables and within-family predictive effects.

@@

Responsiveness and inductive reasoning were measured using the warmth (e.g., "How often do you express affection by hugging, kissing and holding this child?") and inductive reasoning (e.g., "Talk it over and reason with this child when he/she misbehaved?") subscales of the Child Rearing Questionnaire [CRQ, @patersonAssociationBehaviouralAdjustment2001]. Structure and negative parenting were measured using the consistency (e.g., "How often is this child able to get out of punishment when he/she really sets his/her mind to it?", reverse coded) and anger (e.g., "How often are you angry when you punish this child?") scales developed by the Canadian National Longitudinal Survey of Children and Youth [NLSCY; @statisticscanadaNationalLongitudinalSurvey2000]. These self-report instruments used five-point scales, ranging from "never or almost never" to "always or almost always".

@@

With regard to within-families parenting effects, none of the intercepts differed significantly from zero, excepting the effect of maternal warmth on children's emotional adjustment (Table \@ref(tab:tabcrosspath)). Thus, on average, mothers' warmth had a positive significant predictive effect on children's emotional adjustment. The variances of all parenting effects were significant, however, indicating that the effect of qualitative parenting behaviors on children's emotional adjustment varies across families. Moreover, these variances were notably larger for fathers than for mothers, indicating that fathers have a more variable effect on their children through specific parenting behaviors than mothers do.

```{r tabcrosspath, results = "asis"}
tabparenteff <- rbind(tab_std_sections$`Intercepts of parenting effects`, 
                      tab_std_sections$`Between-family level residual variances`[c(1, 5:6),])

@@

tabchildeff <- rbind(tab_std_sections$`Intercepts of child effects`,
                      tab_std_sections$`Between-family level residual variances`[c(1, 7:8),])

tabchildeff$Parameter <- gsub("^\\((.+) on.*$", "\\1", tabchildeff$Parameter)
tabchildeff$Parameter[c(1,4)] <- c("Random intercepts of child effects on:", "Variance of child effects on:")
tab <- rbind(tabparenteff, tabchildeff)
#tab[is.na(tab)] <- ""
rownames(tab) <- NULL
reg_table(tab, caption = "Between-family level intercepts and variances of within-family predictive effects", note = "Est.: Estimate. HDPI: Highest posterior density interval. Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate.")
```

With regard to child effects, in line with hypotheses, the intercept of the cross-lagged effect of emotional adjustment on maternal warmth was positive and significant (Table \@ref(tab:tabcrosspath)). Thus, when children displayed better emotional adjustment than usual in one year, their mothers displayed more anger and inductive reasoning, but also more warmth, than usual in the next year. Contrary to hypotheses, the intercept of the effect of emotional adjustment on maternal consistency was non-significant. As hypothesized, the intercept for maternal inductive reasoning was positive. Finally, counter to hypotheses, the intercept for maternal anger was positive, not negative. In partial support of hypotheses, the intercept for the child effect of adjustment on paternal warmth was positive and significant. None of the other intercepts for child effects on paternal parenting were significant. The variances of all child effects were significant, indicating that the effect of children's emotional adjustment on their parents' child-rearing behaviors differs between families.

@@

To examine interactions between involvement quality and quantity, we tested whether between-family differences in quantitative involvement predicted the within-family cross-lagged effects between qualitative parenting behaviors and emotional adjustment (Table \@ref(tab:tabinvolvement)). With regard to the moderation of parenting effects, the only significant effects we found were for maternal warmth. Specifically, the within-families predictive effect of maternal warmth on emotional adjustment was negatively attenuated when both parents' joint quantitative involvement was higher. This same effect was positively attenuated, however, if mothers' involvement was relative higher than fathers' involvement. We found no moderation for paternal warmth, nor for both parents' consistency, inductive reasoning, and anger.

With regard to the moderation of child effects by quantitative involvement, we found that the effect of adjustment on maternal warmth was negatively attenuated when parents' joint quantitative involvement was higher, and also when mothers' relative involvement was greater compared to fathers. We found no effects for consistency. For inductive reasoning, we found that that the cross-lagged effect of adjustment on maternal inductive reasoning was negatively attenuated when joint involvement was higher. Adjustment was also more negatively predictive of fathers' inductive reasoning when mothers were relatively more involved than fathers. Thus, in families where mothers were more involved than fathers, children's increased emotional adjustment more strongly predicted decreased paternal reasoning - and conversely, decreased emotional adjustment predicted increased paternal reasoning. We found no moderating effects for anger.

>>>>

abstract: |
  The question of whether fathers and mothers play similar roles in children’s
  emotional adjustment has been addressed using either qualitative parenting 
  measures or  quantitative child care involvement. This
  longitudinal study (N = 1884, age 6-12) bridges these approaches, 
  investigating qualitative and quantitative involvement, and their interaction,
  in relation to emotional adjustment. On average, evidence for mothers'
  and fathers' unique roles was limited.
  When both parents were more involved, both engaged
  more in desirable qualitative parenting. When mothers were disproportionately
  involved, however, parenting roles were more differentiated:
  Fathers displayed less
  desirable qualitative parenting, and children were less well-adjusted.
  This suggests that all family members benefit from a more equal
  childcare distribution.

@@

\vspace{\baselineskip}
Emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for the child), and in terms of qualitative parenting behaviors (how do parents behave towards children). Nevertheless, the existing literature has two major shortcomings. Firstly, in assessing linkages with children’s emotional adjustment, quantitative involvement and qualitative parenting behaviors are rarely jointly investigated. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. Children’s emotional adjustment might be differently impacted by parents’  control if that parent is generally highly involved with the day-to-day minutiae of childrearing, than when (s)he is generally absent. Conversely, children’s emotional adjustment might elicit different qualitative parenting behaviors from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be sanctioned. The present study set out to address this question.

Secondly, although fathers have been theorized to play a unique role in emotional adjustment development, empirical work has rarely addressed the joint and relative contributions of fathers' and mothers' involvement. The present study overcomes these limitations by incorporating both quantitative and quantitative measures of mothers' and fathers' involvement. We use sophisticated methods to examine how joint and relative quantitative involvement moderate the longitudinal interplay between mothers' and fathers' qualitative parenting and children's emotional adjustment, from the age of 6 to 12 years. We thus obtain a more comprehensive and nuanced understanding of how fathers' and mothers' involvement relate to children's emotional adjustment development.

@@

Attachment relationships are grounded in a balance between security and exploration [@grossmannDeveloppementAttachementAdaptation1998]. Fatherhood researchers have posited that warmth and security are more central to mother-child attachment relationships, whereas father-child relationships have been construed as "activation relationships", with greater emphasis on exploration [@paquetteFatherChildActivationRelationship2013]. By encouraging children to venture outside their comfort zone, fathers stimulate a broad spectrum of emotions, thereby expanding children's emotional repertoire, and engaging their developing regulatory abilities  [@flouriFathersBehaviorsChildren2010; @bogelsDoesFatherKnow2011; @parkeFatherInvolvement2000]. Consequently, fathers and mothers are theorized to influence emotional adjustment development in somewhat different ways.

@@

It is now widely acknowledged that quantitative and qualitative involvement are both relevant to child development [@faganShouldResearchersConceptualize2014; @pleckPaternalInvolvementRevised2004]. Nevertheless, these dimensions are rarely investigated together, let alone in interaction [@craigDoesFatherCare2006; @faganShouldResearchersConceptualize2014]. The present study overcomes this limitation by investigating both qualitative and quantitative involvement, and the interplay between them, in relation to children's emotional adjustment.

@@

Early developmental research focused on parental socialization; the notion that parents shape child development through proximal mechanisms, such as qualitative parenting behaviors [@maccobySocializationContextFamily1983; see also @vanlissaRoleFathersMothers2018a]. Qualitative parenting behaviors have been operationalized in terms of the dimensions underlying parenting styles: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Substantial evidence indicates that responsiveness, which we operationalized in terms of warm and supportive parenting, is positively associated with children's socio-emotional adjustment [@khalequePerceivedParentalAcceptanceRejection2002; @morrisRoleFamilyContext2007].

The literature linking demandingness and emotional adjustment is less conclusive [@morrisRoleFamilyContext2007]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Some of these might help children meet societal expectations, whereas others are overly restrictive, and interfere with children's autonomy needs [@barberCentralityControlParenting2013]. We examined two aspects of demandingness: Consistency; the extent to which consequences for transgressions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. Consistency has been found to correlate strongly with emotional adjustment [@brandPerceivedParentalInconsistency1990]. There is less empirical evidence for linkages between inductive reasoning and children’s emotional adjustment. Given that inductive inductive reasoning promotes the internalization of parental expectations, including display rules for emotions [@grusecImpactParentalDiscipline1994], we hypothesized that it would be positively related to emotional adjustment.

Negative parenting, which we operationalized as parental anger, is an additional qualitative parenting dimension that is negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. 
Parental anger directly undermines children's emotional adjustment [@scaramellaParentingChildrenDistress2008], and models emotional dysregulation [@changHarshParentingRelation2003]. We thus hypothesized that it would be negatively related to emotional adjustment.

Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' qualitative involvement [@flouriFathersBehaviorsChildren2010]. Although some research suggests that both parents influence children similarly [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique contributions of mothers and fathers. For various reasons, mothers and fathers sometimes appear to fulfill somewhat specialized roles [@bogelsDoesFatherKnow2011]. In line with the theoretical notions of "attachment" and "activation" relationships, mothers tend to encourage, and respond more supportively to, emotional expression, whereas fathers structure and sanction emotion expressions to a greater extent [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, for adolescents' emotional adjustment, maternal warmth was found to be a more salient socialization mechanism than demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. Similar differentiated associations were also found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is some evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is incongruous with the notion of the maternal "attachment" relationship. We might therefore expect emotional adjustment to be associated more strongly with maternal warmth and anger, and with paternal demanding behaviors, such as consistency and inductive reasoning.

### Parent or child effects?

A view of the family as a dynamic system, whose members reciprocally influence one another, has become commonplace [@kerrPerceivedParentingStyle2012]. Nevertheless, bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. The available evidence suggests that maladjustment elicits less parental warmth, more demandingness, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004; @nepplIntergenerationalContinuityParenting2009; @lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional support that emotional maladjustment predicts decreased parental support, and greater demandingness [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014; @wijsbroekDirectionEffectsPerceived2011]. Thus, parents' display punitive responses to maladjustment, including decreased support and increased negative parenting, and corrective responses, such as increased demandingness. We therefore hypothesize that emotional adjustment will positively predict parental warmth, and negatively predict consistency, inductive reasoning, and anger.

### Analyzing the direction of effects

@@

<!-- Hier iets over de hypotheses; dat we niet weten wat er within en wat between speelt -->

@@

Quantitative involvement has been researched extensively in terms of the absolute amount of childcare provided [see @pleckPaternalInvolvementRevised2004]. However, some have argued that the **division** of childcare tasks is more relevant for children's adjustment [@krollEarlyFatherInvolvement2016; @sabattiniRelationMothersFathers2004; @craigHowMothersFathers2011; @feinbergInternalStructureEcological2003]. Indeed, the equity of the childcare distribution is associated with socio-emotional development [@deutschPaternalParticipationChild2001]. Moreover, an unequal distribution often implies that women perform a disproportionate share of the work [@craigDoesFatherCare2006]. This strains the marital relationship [@shockleyItNotWhat2018], and erodes parenting quality, particularly for parents of difficult children [@sheltonInterparentalConflictNegative2008].

There are also methodological advantages to focusing on the distribution of childcare. Operationalizing quantitative involvement in terms of both parents' absolute involvement incurs a potential confound: Effects that appear to be due to fathers' greater involvement might be due to children receiving **more** total care [see @deutschPaternalParticipationChild2001]. This confound can be mitigated by examining the effect of the distribution of childcare, while controlling for total care. Some prior research has attempted to achieve this by assessing the childcare distribution directly, by asking which parent is most responsible for certain tasks [@krollEarlyFatherInvolvement2016; @keizerInfluenceFathersMothers2019].

Alternatively, when quantitative involvement has been measured in terms of absolute involvement, one can estimate a latent mean and difference to statistically disentangle joint involvement from relative involvement [@mcardleLatentVariableModeling2009]. Joint involvement reflects the total (average) amount of care provided to children, and relative involvement reflects the extent to which the mother is more involved than the father (i.e., the childcare distribution).

@@

The amount of time parents spend with children is associated with the quality of parenting behaviors  [@sabattiniRelationMothersFathers2004]. Indeed, more involved parents tend to experience greater parenting self-efficacy, possibly reflecting the effects of experience [@gialloMothersFathersInvolvement2013a]. Furthermore, when both parents are more involved, they monitor each other's parenting behaviors to ensure that children are raised well [@mclanahanConsequencesFatherAbsence2014]. Consequently, we expect greater joint involvement to positively predict both parents' parenting quality.

With regard to differential involvement, we might also expect the more involved parent to display more desirable qualitative parenting behaviors, due to the effects of experience [@sabattiniRelationMothersFathers2004]. However, the benefit of parents monitoring each other's child-rearing should be diminished when parents are not equally involved [@mclanahanConsequencesFatherAbsence2014]. Indeed, retrospective reports indicate that an unequal childcare distribution is associated with suboptimal parenting behaviors: Increased authoritarian parenting in mothers, and increased disengaged parenting in fathers [@sabattiniRelationMothersFathers2004]. Additionally, greater differential involvement might undermine both parents' involvement quality, through its effect on the marital relationship. Childcare responsibilities are a heavy burden [@deutschPaternalParticipationChild2001]. An unfair distribution strains the interparental relationship, interferes with sensitive parenting, and undermines child well-being [@davies2006interparental]. When care is distributed relatively equally, mothers tend to be more satisfied [@pharesFathersMothersInvolvement2009], and the family system more harmonious. Consequently, balanced co-parenting is associated with greater parental warmth [@barnettInterdependenceParentingMothers2008]. We thus expect greater differential involvement to be negatively associated with both parents' parenting quality, and with children's emotional adjustment.

@@

Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. This implies that joint care should be positively associated with children's emotional adjustment. Relative involvement, on the other hand, might be negatively linked to emotional adjustment, because children feel less supported when they perceive that one of their parents is relatively uninvolved, than when both parents are equally invested in their well-being [@deutschPaternalParticipationChild2001].

@@

Interactions between involvement quality and quantity have largely been overlooked in the parenting literature [@jaffeeLifeFatherBenefits2003]. One notable exception is Feinberg's coparenting model [@feinbergInternalStructureEcological2003]. Coparenting refers to the ways parents relate to each other in the parenting role. Although the coparenting literature is typically invoked to describe child-rearing in divorced families, it also holds insights for intact families. According to Feinberg, the "division of labor" in childcare, analogous to our operationalization of relative involvement, affects qualitative child-rearing behaviors and children's adjustment directly, and moderates the link between these two constructs.

@@

Sex differences are a relevant factor in the interplay between parenting and emotional adjustment. Parents tend to raise boys and girls differently; expressing more demandingness with girls [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2018a], and more anger with boys [@strausCorporalPunishmentAmerican1999]. Furthermore, social learning theory posits that the same-sex parent plays a stronger role in socialization; particularly for sex-typed behavior, such as emotion display rules [@busseySocialCognitiveTheory1999]. Finally, mounting evidence suggests that parental warmth is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2018a]. We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning towards girls, and more anger towards boys. Furthermore, we predict stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent.

Furthermore, although sex differences in emotional adjustment are sometimes found in older children [@silkAdolescentsEmotionRegulation2003; @vanlissaRoleFathersMothers2018a], a meta-analysis found little evidence for such sex differences in young children [@else-questGenderDifferencesTemperament2006a]. Consequently, we have no hypothesis regarding sex differences in children's average levels of emotional adjustment.

@@

The present study aims to make several contributions. First, we include both mothers and fathers, in order to tease apart their common and unique roles in children's emotional adjustment development. Second, we incorporate both qualitative and quantitative measures of involvement, and examine interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work. Third, we disentangle stable between-family differences from dynamic within-family predictive processes, in order to more accurately assess the direction of effects between qualitative involvement and emotional adjustment. Finally, we examined sex as a moderator of between-family differences in all study variables and within-family predictive effects.

@@

Responsiveness and inductive reasoning were measured using the warmth (e.g., "How often do you express affection by hugging, kissing and holding this child?") and inductive reasoning (e.g., "Talk it over and reason with this child when he/she misbehaved?") subscales of the Child Rearing Questionnaire [CRQ, @patersonAssociationBehaviouralAdjustment2001]. Consistency and negative parenting were measured using the consistency (e.g., "How often is this child able to get out of punishment when he/she really sets his/her mind to it?", reverse coded) and anger (e.g., "How often are you angry when you punish this child?") scales developed by the Canadian National Longitudinal Survey of Children and Youth [NLSCY; @statisticscanadaNationalLongitudinalSurvey2000]. These self-report instruments used five-point scales, ranging from "never or almost never" to "always or almost always".

@@

With regard to within-families parenting effects, none of the intercepts differed significantly from zero, excepting the effect of maternal warmth on children's emotional adjustment (Table \@ref(tab:tabcrosspath)). Thus, on average, mothers' warmth had a positive significant predictive effect on children's emotional adjustment. The variances of all parenting effects were all significant, however, indicating that the effect of qualitative parenting behaviors on children's emotional adjustment varies across families. Moreover, these variances were notably larger for fathers than for mothers, indicating that fathers have a more variable effect on their children through specific parenting behaviors than mothers do.

```{r tabcrosspath, results = "asis"}
tabparenteff <- rbind(tab_std_sections$`Intercepts of parenting effects`, 
                      tab_unst_sections$`Between-family level residual variances`[c(1, 5:6),])

@@

tabchildeff <- rbind(tab_std_sections$`Intercepts of child effects`,
                      tab_unst_sections$`Between-family level residual variances`[c(1, 7:8),])

tabchildeff$Parameter <- gsub("^\\((.+) on.*$", "\\1", tabchildeff$Parameter)
tabchildeff$Parameter[c(1,4)] <- c("Random intercepts of child effects on:", "Variance of child effects on:")
tab <- rbind(tabparenteff, tabchildeff)
#tab[is.na(tab)] <- ""
rownames(tab) <- NULL
tab$Parameter <- gsub("^Variance", "Unstandardized variance", tab$Parameter)
reg_table(tab, caption = "Standardized between-family level intercepts and unstandardized variances of within-family predictive effects", note = "Est.: Estimate. HDPI: Highest posterior density interval. Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate.")
```

With regard to child effects, in line with hypotheses, the intercept of the cross-lagged effect of emotional adjustment on maternal warmth was positive and significant (Table \@ref(tab:tabcrosspath)). Thus, when children displayed better emotional adjustment than usual in one year, their mothers displayed more warmth than usual in the next year. Contrary to hypotheses, the intercept of the effect of emotional adjustment on maternal consistency was non-significant. As hypothesized, the intercept for maternal inductive reasoning was positive. Counter to hypotheses, the intercept for maternal anger was positive, not negative. Thus, when children displayed better emotional adjustment than usual in one year, their mothers displayed more anger than usual in the next year. In partial support of hypotheses, the intercept for the child effect of adjustment on paternal warmth was positive and significant. None of the other intercepts for child effects on paternal parenting were significant. The variances of all child effects were all significant but small, indicating that the effect of children's emotional adjustment on their parents' child-rearing behaviors differs somewhat between families.

@@

To examine interactions between involvement quality and quantity, we tested whether between-family differences in quantitative involvement predicted the within-family cross-lagged effects between qualitative parenting behaviors and emotional adjustment (Table \@ref(tab:tabinvolvement)). With regard to the moderation of parenting effects, the only significant effects we found were for maternal warmth. Specifically, the within-families predictive effect of maternal warmth on emotional adjustment was negatively attenuated when joint involvement was higher, and positively attenuated when mothers were relatively more involved than fathers. We found no moderation for paternal warmth, nor for both parents' consistency, inductive reasoning, and anger.

With regard to the moderation of child effects by quantitative involvement, we found that the effect of adjustment on maternal warmth was negatively attenuated when parents' joint quantitative involvement was higher, and also when mothers' relative involvement was greater compared to fathers. We found no effects for consistency. For inductive reasoning, we found that that the cross-lagged effect of adjustment on maternal inductive reasoning was negatively attenuated when joint involvement was higher. Emotional adjustment was also more negatively predictive of fathers' inductive reasoning when mothers were relatively more involved than fathers. Thus, in families where mothers were more involved than fathers, children's increased emotional adjustment more strongly predicted decreased paternal reasoning - and conversely, decreased emotional adjustment predicted increased paternal reasoning. We found no moderating effects for anger.
